---
title: "Prepare VAERS data for safety signal detection"
author:
  - name: Nan Xiao
    url: https://nanx.me/
    affiliation: Seven Bridges
    affiliation_url: https://www.sevenbridges.com/
  - name: Soner Koc
    url: https://github.com/skoc
    affiliation: Seven Bridges
    affiliation_url: https://www.sevenbridges.com/
  - name: Kaushik Ghose
    url: https://kaushikghose.wordpress.com/
    affiliation: Seven Bridges
    affiliation_url: https://www.sevenbridges.com/
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This page describes the steps involved to downloaded and preprocess the raw VAERS data, and transform the data into an analyzable format.

# Download the data

To download the VAERS data:

- Go to the [VAERS Data Sets page](https://vaers.hhs.gov/data/datasets.html).
- Download the zip files from 1990 to 2019 (need to pass the captchas).
- Put all downloaded zip files in the `data/` folder as is.

As time of the downloading (Feburary 2020), data from all years since 1990
was complete except for 2019. The data for 2019 contains VAERS
reports processed as of 2019-12-14.

The captchas on each dataset download page might make it a bit complicated
to automate the data retrieval process, especially when 30 years of
data needs to be downloaded. Providing an standard, programatically
accessible API for retriving the data like the current AERS database
would be helpful.

# Load the data

```{r}
library("readr")
library("dplyr")
library("tidyr")

year <- as.character(1990:2019)
k <- length(year)
file_zip <- paste0("data/", year, "VAERSData.zip")
file_meta <- paste0(year, "VAERSDATA.csv")
file_vax <- paste0(year, "VAERSVAX.csv")
file_sym <- paste0(year, "VAERSSYMPTOMS.csv")

lst_meta <- vector("list", k)
lst_vax <- vector("list", k)
lst_sym <- vector("list", k)

for (i in 1:k) {
  cat("Reading VAERS data from year", year[i], "\n")
  con1 <- unz(description = file_zip[i], filename = file_meta[i])
  lst_meta[[i]] <- read_csv(con1, col_types = cols(VAERS_ID = col_character()))

  con2 <- unz(description = file_zip[i], filename = file_vax[i])
  lst_vax[[i]] <- read_csv(con2, col_types = cols(VAERS_ID = col_character()))

  con3 <- unz(description = file_zip[i], filename = file_sym[i])
  lst_sym[[i]] <- read_csv(con3, col_types = cols(VAERS_ID = col_character()))
}

df_meta <- dplyr::bind_rows(lst_meta)
df_vax <- dplyr::bind_rows(lst_vax)
df_sym <- dplyr::bind_rows(lst_sym)

df_meta <- df_meta[, c("VAERS_ID", "AGE_YRS", "SEX")]
df_vax <- df_vax[, c("VAERS_ID", "VAX_NAME")]
df_sym <- df_sym[, c("VAERS_ID", "SYMPTOM1", "SYMPTOM2", "SYMPTOM3", "SYMPTOM4", "SYMPTOM5")]
```

# Quality Control

To ensure the basic data quality, remove all rows with `NA`s from the metadata table:

```{r}
df_meta <- df_meta[complete.cases(df_meta), ]
```

Concatenate columns for symptoms:

```{r}
df_sym$"SYMPTOM" <- paste(df_sym$SYMPTOM1, df_sym$SYMPTOM2, df_sym$SYMPTOM3, df_sym$SYMPTOM4, df_sym$SYMPTOM5, sep = ", ")
```

Use the VAERS IDs that are common across all tables as the index:

```{r}
id <- intersect(intersect(unique(df_meta$VAERS_ID), unique(df_vax$VAERS_ID)), unique(df_sym$VAERS_ID))
```

Keep only the VAERS IDs shared across all tables:

```{r}
df_meta <- df_meta[df_meta$VAERS_ID %in% id, ]
df_vax <- df_vax[df_vax$VAERS_ID %in% id, ]
df_sym <- df_sym[df_sym$VAERS_ID %in% id, ]
```

Squash the same IDs but multiple symptoms or multiple vaccines into one row:

```{r}
df_vax <- df_vax %>%
  group_by(VAERS_ID) %>%
  arrange(VAX_NAME) %>%
  summarise(VAX_NAMES = paste(VAX_NAME, collapse = ", "))

df_sym <- df_sym %>%
  group_by(VAERS_ID) %>%
  arrange(SYMPTOM) %>%
  summarise(SYMPTOMS = paste(SYMPTOM, collapse = ", "))
```

# Merge tables

Merge all tables

```{r}
df <- inner_join(inner_join(df_meta, df_vax, by = "VAERS_ID"), df_sym, by = "VAERS_ID")
```

Separate the rows with multiple vaccines or multiple symptoms into multiple rows:

```{r}
df <- separate_rows(df, VAX_NAMES, sep = ", ")
df <- separate_rows(df, SYMPTOMS, sep = ", ")
df <- df[(df$SYMPTOMS != "NA"), ]
```

# Recode variables

Recode stratification variables and recode age groups:

```{r}
names(df) <- c("id", "strat_age", "strat_gender", "var1", "var2")
df$strat_gender <- recode(df$strat_gender, F = "Female", M = "Male", U = "Unknown")

idx1 <- which(df$strat_age <= 2)
idx2 <- which(df$strat_age > 2 & df$strat_age <= 18)
idx3 <- which(df$strat_age > 18 & df$strat_age <= 65)
idx4 <- which(df$strat_age > 65)

df$strat_age[idx1] <- "<= 2"
df$strat_age[idx2] <- "2 - 18"
df$strat_age[idx3] <- "18 - 65"
df$strat_age[idx4] <- "> 65"

df$strat_gender <- as.factor(df$strat_gender)
df$strat_age <- as.factor(df$strat_age)
```

# Save to file

Rearrange variable order:

```{r}
df <- df[, c("id", "var1", "var2", "strat_age", "strat_gender")]
```

Sanity check:

```{r}
dim(df)
```

```
# [1] 3441506       5
```

```{r}
head(df)
```

```
# # A tibble: 6 x 5
#   id     var1                var2        strat_age strat_gender
#   <chr>  <chr>               <chr>       <fct>     <fct>       
# 1 025001 DTP (NO BRAND NAME) Agitation   <= 2      Female      
# 2 025003 DTP (TRI-IMMUNOL)   Delirium    <= 2      Male        
# 3 025003 DTP (TRI-IMMUNOL)   Hypokinesia <= 2      Male        
# 4 025003 DTP (TRI-IMMUNOL)   Hypotonia   <= 2      Male        
# 5 025003 POLIO VIRUS         Delirium    <= 2      Male        
# 6 025003 POLIO VIRUS         Hypokinesia <= 2      Male      
```

Save to file:

```{r}
saveRDS(df, file = "data-processed/df.rds")
```
